\section{Strategic Recommendations}
\label{sec:recommendations}

\tinypara{Priority 1: Implement Comprehensive \gls{c-scrm} Program.} Establish formal programs per \gls{nist} \gls{sp} \href{https://csrc.nist.gov/pubs/sp/800/161/r1/upd1/final}{800-161 Rev. 1} \cite{boyens2024}: maintain comprehensive SBOM, conduct supplier risk assessments, establish contractual security requirements, and implement continuous supply chain monitoring.

\tinypara{Priority 2: Adopt Zero Trust Architecture.} Implement Zero Trust per \gls{nist} \gls{sp} \href{https://csrc.nist.gov/pubs/sp/800/207/final}{800-207} \cite{rose2020}: network micro-segmentation preventing lateral movement, universal \gls{mfa} with hardware tokens, least privilege access, and comprehensive device trust verification.

\tinypara{Priority 3: Build Integrity \& Comprehensive Detection.} Ensure compiled artifacts match source code through reproducible builds from multiple independent pipelines with cryptographic verification. Deploy monitoring across all environments, including universal \gls{edr}, \gls{siem} integration of development logs, and \gls{ueba} for privileged accounts \cite{litan2015}.

\tinypara{Priority 4: Elevate Governance \& Foster Security Culture.} Establish board-level oversight with cybersecurity expertise, ensuring supply chain risks are a regular agenda item. Foster a security-first culture through regular training and security champions programs.

\subsection{Implementation Roadmap}

\tinypara{Immediate (0-3 months):} MFA enforcement, EDR deployment to development systems, emergency \gls{c-scrm} assessment, board briefing.

\tinypara{Short-term (3-6 months):} \gls{sbom} implementation, build integrity verification, network micro-segmentation, \gls{ueba} deployment.

\tinypara{Medium-term (6-12 months):} Complete Zero Trust architecture, comprehensive supplier assessments, security culture programs.

\tinypara{Long-term (12+ months):} Continuous improvement through threat intelligence integration, regular exercises, security maturity progression.
