% Section 5: Discussion
\section{Discussion and Best Practices}

\subsection{When to Use Prompt-Based APIs}

Prompt-based \gls{api}s excel where natural language understanding provides genuine value: complex query interpretation (search, recommendations, assistants), content generation (summarisation, translation, code generation), and multi-step reasoning requiring sequential decision-making through techniques like chain-of-thought prompting \cite{brown_language_2020, schluntz_building_2024, wei_chainofthought_2022}.

\textbf{When REST Remains Superior:} \gls{crud} operations, high-frequency transactions, and deterministic responses should retain REST architecture. Prompt-based approaches introduce latency ($100-2000ms$ vs. $<10ms$), non-determinism, and cost implications (\$0.077 per 1K requests for Nova Micro vs. negligible for \gls{rest}) \cite{aws_bedrock_2025}.

\textbf{Service-Oriented Architecture Evolution:} Prompt-based \gls{api}s represent service-oriented evolution rather than replacement. \gls{soap} emphasized formal contracts (\gls{wsdl}), \gls{rest} emphasized resource design (\gls{http}), and prompt-based designs emphasize intent (natural language). Modern architectures should combine paradigms: \gls{rest} for transactions, prompt-based for natural language \cite{lercher_managing_2024}.

\subsection{Design Recommendations}

\textbf{Model Selection:} Select the smallest model capable of reliable execution. Open-source models such as Llama~2 \cite{touvron_llama_2023} suit data-sovereignty constraints but require significant infrastructure overhead; for managed deployments, Nova Micro achieves 83\% cost savings versus Llama 3 8B \cite{aws_nova_2025}.

\textbf{Prompt Engineering:} Include explicit output formats, few-shot examples (2-3 samples), and role-based framing \cite{chen_prompt_2025}. Separate system instructions from user content using structural markers to prevent prompt injection \cite{owasp_top_2024, schluntz_building_2024}.

\textbf{Output Validation:} Implement schema validation, semantic guardrails (\gls{aws} \href{https://aws.amazon.com/bedrock/guardrails/}{Bedrock Guardrails}), and fallback mechanisms (retry, rule-based processing, escalation) \cite{lercher_managing_2024, aws_bedrock_2025}.

\textbf{Versioning and Cost:} Store prompt templates in version control, test changes, maintain multiple versions. For optimization: remove verbosity, leverage caching, aggregate requests, route simple tasks to smaller models \cite{sahoo_systematic_2025}.

\subsection{Future Directions}

Multi-agent orchestration addresses complex workflows through specialized \gls{llm} collaboration \cite{schluntz_building_2024}. Retrieval-augmented generation mitigates hallucination \cite{zhou_hademif_2025}. Standardization efforts could establish common schemas \cite{white_prompt_2023}. As prompt-based \gls{api}s move into safety-critical domains, formal verification will mature \cite{aws_bedrock_2025}. Prompt caching, already supported by providers such as Anthropic and \gls{aws}, will become a standard cost-reduction primitive as token expenses scale with adoption. Multimodal inputs, combining voice, image, and text within a single \gls{api} call, represent the next frontier, dissolving the boundary between conversational and traditional web services entirely. Prompt-based \gls{api}s will complement \gls{rest} in hybrid architectures. The challenge lies in thoughtfully architecting systems that harness both paradigms' strengths.
