\chapter{Data Preparation}

The datasets used in this study are MedQA-\gls{usmle} \cite{jin_what_2021} and MedMCQA \cite{pal_medmcqa_2022}, both of which are also used for benchmark evaluation in the \gls{amg-rag} methodology of \citet{rezaei_agentic_2025}. MedQA is a multiple-choice question answering dataset derived from professional medical licensing examinations \cite{jin_what_2021}, and its questions were written in reference to a collection of 18 standard medical textbooks. MedMCQA is built from medical entrance examination data and covers both foundational and clinical knowledge across a broad range of medical specialties \cite{pal_medmcqa_2022}. 

For the purposes of this system, the MedQA textbooks serve as the \gls{rag} retrieval corpus, while MedMCQA is used exclusively to train the specialty classifier, as it provides explicit subject labels that MedQA lacks. Initial exploratory analysis confirmed answer balance in MedQA (uniform distribution across options A-D) and characterised the specialty distribution in MedMCQA (See Figure \ref{fig:textbooks} in Appendix). Textbook character counts and chunk distributions were subsequently verified prior to vectorstore construction.

Following these exploratory checks, the MedQA textbooks are chunked into 400-word sections with a 50-word overlap, producing 36,723 chunks across all 18 books. To support the embedding model ablation described in Chapter \ref{ch:ablation}, three separate \gls{faiss} vectorstores were constructed: one each for all-MiniLM-L12-v2, Qwen3-Embedding-0.6B, and Octen-Embedding-0.6B. Qwen3-Embedding-0.6B was selected as the build default due to its earlier availability: at 0.6B parameters, it fits within the Colab T4 \gls{vram} budget. 

As validated by the ablation grid (Chapter \ref{ch:ablation}), Octen-Embedding-0.6B ultimately outperforms Qwen3-Embedding-0.6B in the \gls{rag} pipeline, delivering a +11\gls{pp} accuracy gain with the Thinking-2507 \gls{llm} against a +1\gls{pp} result for Qwen3-Embedding under the same conditions. The choice of embedding model is therefore a primary experimental variable, and the rationale for both selections is detailed in Section \ref{sec:model_selection}.
